---
title: "R Maps for Visualizing Spatial Information"
author: "Echefu Chinwendu"
date: "2024-03-19"
output:
  html_document: 
    toc: yes
    toc_depth: 4
    toc_float: yes
    fig_width: 4
    fig_caption: yes
    number_sections: yes
    toc_collapsed: yes
    code_folding: hide
    code_download: yes
    smooth_scroll: true
    theme: readable
    fig_height: 4
  pdf_document: 
    toc: yes
    toc_depth: 4
    fig_caption: yes
    number_sections: yes
    fig_width: 3
    fig_height: 3
  word_document: 
    toc: yes
    toc_depth: 4
    fig_caption: yes
    keep_md: yes
editor_options: 
  chunk_output_type: inline
---

```{=html}
<style type="text/css">

/* Table of content - navigation */
div#TOC li {
    list-style:none;
    background-color:lightgray;
    background-image:none;
    background-repeat:none;
    background-position:0;
    font-family: Arial, Helvetica, sans-serif;
    color: #780c0c;
}


/* Title fonts */
h1.title {
  font-size: 24px;
  color: darkblue;
  text-align: center;
  font-family: Arial, Helvetica, sans-serif;
  font-variant-caps: normal;
}
h4.author { 
  font-size: 18px;
  font-family: Arial, Helvetica, sans-serif;
  color: navy;
  text-align: center;
}
h4.date { 
  font-size: 18px;
  font-family: Arial, Helvetica, sans-serif;
  color: darkblue;
  text-align: center;
}

/* Section headers */
h1 {
    font-size: 22px;
    font-family: "Times New Roman", Times, serif;
    color: darkred;
    text-align: left;
}

h2 {
    font-size: 18px;
    font-family: "Times New Roman", Times, serif;
    color: navy;
    text-align: left;
}

h3 { 
    font-size: 15px;
    font-family: "Times New Roman", Times, serif;
    color: darkred;
    text-align: left;
}

h4 {
    font-size: 18px;
    font-family: "Times New Roman", Times, serif;
    color: darkred;
    text-align: left;
}

/* Decoration of hyperlinks  */

/* unvisited link */
a:link {
  color: green;
}

/* visited link */
a:visited {
  color: purple;
}

/* mouse over link */
a:hover {
  color: red;
}

/* selected link */
a:active {
  color: yellow;
}
</style>
```


```{r setup, include=FALSE}
# code chunk specifies whether the R code, warnings, and output 
# will be included in the output files.
options(repos = list(CRAN="http://cran.rstudio.com/"))
if (!require("tidyverse")) {
   install.packages("tidyverse")
   library(tidyverse)
}
if (!require("knitr")) {
   install.packages("knitr")
   library(knitr)
}
if (!require("cowplot")) {
   install.packages("cowplot")
   library(cowplot)
}
if (!require("latex2exp")) {
   install.packages("latex2exp")
   library(latex2exp)
}
if (!require("plotly")) {
   install.packages("plotly")
   library(plotly)
}
if (!require("gapminder")) {
   install.packages("gapminder")
   library(gapminder)
}
if (!require("png")) {
    install.packages("png")             # Install png package
    library("png")
}
if (!require("RCurl")) {
    install.packages("RCurl")             # Install RCurl package
    library("RCurl")
}
if (!require("colourpicker")) {
    install.packages("colourpicker")              
    library("colourpicker")
}
if (!require("gganimate")) {
    install.packages("gganimate")              
    library("gganimate")
}
if (!require("gifski")) {
    install.packages("gifski")              
    library("gifski")
}
if (!require("magick")) {
    install.packages("magick")              
    library("magick")
}
if (!require("grDevices")) {
    install.packages("grDevices")              
    library("grDevices")
}
if (!require("jpeg")) {
    install.packages("jpeg")              
    library("jpeg")
}
if (!require("VGAM")) {
    install.packages("VGAM")              
    library("VGAM")
}
if (!require("MASS")) {
    install.packages("MASS")              
    library("MASS")
}
if (!require("nnet")) {
    install.packages("nnet")              
    library("nnet")
}
if (!require("cluster")) {
    install.packages("cluster")              
    library("cluster")
}
if (!require("stringr")) {
   install.packages("stringr", dependencies = TRUE)
   library(stringr)
}

if (!require("tm")) {
   install.packages("tm", dependencies = TRUE)
   library(tm)
}

if (!require("wordcloud")) {
   install.packages("wordcloud", dependencies = TRUE)
   library(wordcloud)
}

if (!require("RCurl")) {
   install.packages("RCurl", dependencies = TRUE)
   library(RCurl)
}

if (!require("XML")) {
   install.packages("XML", dependencies = TRUE)
   library(XML)
}
if (!require("SnowballC")) {
   install.packages("SnowballC", dependencies = TRUE)
   library(SnowballC)
}
if (!require("RColorBrewer")) {
   install.packages("RColorBrewer", dependencies = TRUE)
   library(RColorBrewer)
}
if (!require("ggplot2")) {
    install.packages("ggplot2")              
    library("ggplot2")
}
if (!require("gganimate")) {
    install.packages("gganimate")              
    library("gganimate")
}
if (!require("graphics")) {
    install.packages("graphics")              
    library("graphics")
}
if (!require("ggridges")) {
    install.packages("ggridges")              
    library("ggridges")
}
if (!require("plyr")) {
    install.packages("plyr")              
    library("plyr")
}
if (!require("ggiraph")) {
    install.packages("ggiraph")              
    library("ggiraph")
}
if (!require("highcharter")) {
    install.packages("highcharter")              
    library("highcharter")
}
if (!require("forecast")) {
    install.packages("forecast")              
    library("forecast")
}
if (!require("leaflet")) {
    install.packages("leaflet")              
    library("leaflet")
}
if (!require("maps")) {
    install.packages("maps")              
    library("maps")
}
if (!require("htmltools")) {
    install.packages("htmltools")              
    library("htmltools")
}
if (!require("htmlwidgets")) {
    install.packages("htmlwidgets")              
    library("htmlwidgets")
}
if (!require("leaflegend")) {
    install.packages("leaflegend")              
    library("leaflegend")
}
if (!require("geojsonio")) {
    install.packages("geojsonio")              
    library("geojsonio")
}
if (!require("stringi")) {
    install.packages("stringi")              
    library("stringi")
}
if (!require("RColorBrewer")) {
    install.packages("RColorBrewer")              
    library("RColorBrewer")
}
if (!require("tigris")) {
    install.packages("tigris")              
    library("tigris")
}
if (!require("leafpop")) {
    install.packages("leafpop")              
    library("leafpop")
}
if (!require("leafem")) {
    install.packages("leafem")              
    library("leafem")
}
if (!require("tmap")) {
    install.packages("tmap")              
    library("tmap")
}
if (!require("tmaptools")) {
    install.packages("tmaptools")              
    library("tmaptools")
}
if (!require("webshot2")) {
    install.packages("webshot2")              
    library("webshot2")
}
if (!require("sf")) {
    install.packages("sf")              
    library("sf")
}
if (!require("terra")) {
    install.packages("terra")              
    library("terra")
}
if (!require("leafpop")) {
    install.packages("leafpop")              
    library("leafpop")
}
if (!require("mapview")) {
    install.packages("mapview")              
    library("mapview")
}
if (!require("spData")) {
    install.packages("spData")              
    library("spData")
}

knitr::opts_chunk$set(echo = TRUE,       
                      warning = FALSE,   
                      result = TRUE,   
                      message = FALSE,
                      comment = NA)
```

# Introduction
Two datasets were used for this project. The first is the gas station dataset. This dataset contains 72798 observations and 31 variables. It gives information about the various locations of gas stations in the US. Some of the variables include state, county, city, longitude, latitude, availability of diesel and so on. 

The second dataset is the Philly Crime Data (2015 - 2024). It is a longitudinal data that contains crime cases in the Philadelphia area since 2015. It has 15243 observations and 19 variables. Some of the variables include: race, sex, date of crime, fatal crimes, street name, block number, zip code and so on.

The datasets can be found here:

Gas Station Dataset: https://github.com/chinwex/STA553/raw/main/w07/POC.csv

Philly Crime Dataset:https://github.com/chinwex/STA553/raw/main/w07/PhillyCrimeSince2015.csv

# Gas Station Data
The table below gives information about the first six gas stations in the dataset.
```{r}
dt <- read.csv(file="https://github.com/chinwex/STA553/raw/main/w07/POC.csv")[,-1]
kable(head(dt))
```

## Map Showing Different Gas Station Locations
A random sample of 500 gas stations were taken from the dataset to plot on the map.
```{r}
set.seed(100)  
dt500 <- dt[sample(nrow(dt), size = 500), ]
```


```{r fig.width=7, fig.height=5, fig.cap = "A Map of 500 Gas Stations in the US"}
title1 <- tags$div( HTML('<font color = "darkred" size =5><b>A Map of 500 Gas Stations in the US</b></font>'))

leaflet(dt500) %>% 
  addTiles() %>%
  setView(lng = -75.5978, lat=39.9522, zoom = 9) %>%
  addMarkers(~xcoord, ~ycoord, popup = ~paste("State: ",STATE, 
                                         "<br>County: ", county,
                                         "<br>Address: ", ADDRESS,
                                         "<br>Zipcode: ", ZIPnew))%>%
 addControl(title1, position = "topright", className = "map-title")
```

# Philly Crime Data
The table below shows the first 6 crime cases in the dataset.
```{r}
phily <- na.omit(read.csv(file = "https://github.com/chinwex/STA553/raw/main/w07/PhillyCrimeSince2015.csv"))
kable(head(phily))
```

## Map Showing the 2023 Philly Crime Data 
All 2023 crime cases where filtered from the dataset and shown in the map below.
```{r}
phily$year <- format(as.Date(phily$date, format = "%m/%d/%Y"), "%Y")
phily2023 <- filter(phily, year == 2023)
```

```{r fig.width=7, fig.height=5, fig.cap = "A Map of Philly Crime Locations in 2023"}

philcolor <- rep("#0072B2", length(phily2023$fatal))
philcolor[which(phily2023$fatal == "Fatal")] <- "#D55E00"

labels <-  paste("Street Name: ",phily2023$street_name, 
             "<br>Block Number: ", phily2023$block_number,
             "<br>Neighborhood: ", phily2023$neighborhood,
             "<br>Zipcode: ", phily2023$zip_code)%>% 
                      lapply(htmltools::HTML)

title <- tags$div( HTML('<font color = "purple" size =5><b>2023 Philly Crime Locations</b></font>'))

 Annotat <- tags$div(HTML('<center><font color = "blue" size =3>The circle sizes are proportional to the ages</font>'))


leaflet(phily2023) %>%
  addTiles() %>% 
  setView(lng = mean(phily2023$lng), lat=mean(phily2023$lat), zoom = 15) %>%
  addCircleMarkers(
            ~lng, 
            ~lat,
            color = philcolor,
            fillColor = ifelse(phily2023$fatal == "Nonfatal", "#ffff66", "#ff99ff"),
            radius = ~(phily2023$age/10)*2,
            opacity = 1,
           # stroke = FALSE, 
           fillOpacity = 0.25,
            label = ~labels ) %>%
  
  addLegend(position = "bottomright", 
            colors = c("#0072B2", "#D55E00"),
            labels= c("Nonfatal", "Fatal"),
            title= "Crime Type",
            opacity = 0.5)%>%
 addControl(title, position = "topright", className = "map-title")%>%
  addControl(Annotat, position = "bottomleft")%>%

addLegendSize(position = 'bottomright', 
                  values = (phily2023$age/10)*2,
                   color = 'gray',
               fillColor = 'gray',
                 opacity = .5,
                   title = 'Age',
                   shape = 'circle',
             orientation = 'vertical',
                  breaks = 4)

```

## Map Interpretation
The map above shows various crime locations in Philadelphia with orange circles representing fatal crimes and blue circles representing nonfatal crimes. From the plot, it is clear that in 2023, there were more nonfatal cases than fatal ones. The circle sizes are proportional to the ages on the dataset. Most of the fatal crimes are associated with younger ages. Majority of crime cases are seen in North Philadelphia.










